### 2016-11-15 greedy_hamiltonian_new_template/paper.tex
fist commit

%%
%% This is file `docultex.tex', 
%% Documentation for siam supplemental file macros for use with LaTeX 2e
%% 
%% December 19, 2013
%%
%% Supplemental Files Supported Version 1.0.0
%% 
%% You are not allowed to change this file. 
%% 
%% You are allowed to distribute this file under the condition that 
%% it is distributed together with all of the files in the siam macro 
%% distribution. These are:
%%
%%  siamltex1213.cls (main LaTeX macro for SIAM)
%%  siam10.clo   (size option for 10pt papers)
%%  subeqn.clo   (allows equation numbners with lettered subelements)
%%  siam.bst     (bibliographic style file for BibTeX)
%%  docultex.tex (this file)
%%
%% If you receive only some of these files from someone, complain! 
%% 
%% You are NOT ALLOWED to distribute this file alone. You are NOT 
%% ALLOWED to take money for the distribution or use of either this 
%% file or a changed version, except for a nominal charge for copying 
%% etc. 
%% \CharacterTable
%%  {Upper-case    \A\B\C\D\E\F\G\H\I\J\K\L\M\N\O\P\Q\R\S\T\U\V\W\X\Y\Z
%%   Lower-case    \a\b\c\d\e\f\g\h\i\j\k\l\m\n\o\p\q\r\s\t\u\v\w\x\y\z
%%   Digits        \0\1\2\3\4\5\6\7\8\9
%%   Exclamation   \!     Double quote  \"     Hash (number) \#
%%   Dollar        \$     Percent       \%     Ampersand     \&
%%   Acute accent  \'     Left paren    \(     Right paren   \)
%%   Asterisk      \*     Plus          \+     Comma         \,
%%   Minus         \-     Point         \.     Solidus       \/
%%   Colon         \:     Semicolon     \;     Less than     \<
%%   Equals        \=     Greater than  \>     Question mark \?
%%   Commercial at \@     Left bracket  \[     Backslash     \\
%%   Right bracket \]     Circumflex    \^     Underscore    \_
%%   Grave accent  \`     Left brace    \{     Vertical bar  \|
%%   Right brace   \}     Tilde         \~}

\documentclass[final,leqno,onefignum,onetabnum]{siamltex1213}

%-------------------------------------------------------------------------
%
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%
%-------------------------------------------------------------------------

\usepackage{./pkg/packages}

\title{Structure Preserving Model Reduction of Parametric Hamiltonian Systems}

\author{
	Babak Maboudi Afkham\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (babak.maboudi@epfl.ch)} 
	\and 
	Jan S. Hesthaven\thanks{Department of Mathematics, Chair of Computational Mathematics and Simulation Schience (MCSS), \'Ecole Polytechnique F\'ed\'erale de Lausanne, Switzerland (jan.hesthaven@epfl.ch)}}

\begin{document}
\maketitle
\slugger{mms}{xxxx}{xx}{x}{x--x}%slugger should be set to mms, siap, sicomp, sicon, sidma, sima, simax, sinum, siopt, sisc, or sirev

\begin{abstract}
While reduced-order models (ROMs) are popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question. We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model. Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the reduce basis. We used the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors. We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters. Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure. This enables the reduction of computational cost for nonlinear Hamiltonian systems. The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.

%Through the greedy selection of basis vectors, we used the Hamiltonian function as an error estimator to search for the

%to add basis vectors, two at a time, and increase the overall accuracy of the reduced basis. It therefor requires a single numerical solution to be computed at each iteration which can save computational costs in stage offline stage. Under

%We also illustrated that convergence rate of the greedy approach is proportional to the best possible convergence rate of any possible iterative update of the reduced basis. Moreover, for nonlinear Hamiltonian systems, we combined the reduced basis obtained from the greedy method with discrete empirical interpolation method and demonstrated the

\end{abstract}

\begin{keywords}Symplectic model reduction, Hamiltonian system, Greedy basis generation, Symplectic Discrete Empirical Interpolation (SDEIM) \end{keywords}

\begin{AMS}\end{AMS}

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{Babak Maboudi Afkham and Jan S. Hesthaven}{Symplectic Model reduction of Parametric Hamiltonian Systems}

@@

\input{./chapters/appendix.tex}

\bibliography{ref}
\bibliographystyle{plain}

\end{document}
%% end of file `docultex.tex'

>>>>

% SIAM Article Template
\documentclass[review]{siamart}
\usepackage{mwe}
\usepackage{subfig}

% Information that is shared between the article and the supplement
% (title and author information, macros, packages, etc.) goes into
% ex_shared.tex. If there is no supplement, this file can be included
% directly.

\input{ex_shared}

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={\TheTitle},
  pdfauthor={\TheAuthors}
}
\fi

% The next statement enables references to information in the
% supplement. See the xr-hyperref package for details.

\externaldocument{ex_supplement}

% FundRef data to be entered by SIAM
%<funding-group>
%<award-group>
%<funding-source>
%<named-content content-type="funder-name"> 
%</named-content> 
%<named-content content-type="funder-identifier"> 
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>

\begin{document}

\maketitle

% REQUIRED
\begin{abstract}
While reduced-order models (ROMs) are popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question. We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model. Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the reduce basis. We used the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors. We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters. Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure. This enables the reduction of computational cost for nonlinear Hamiltonian systems. The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.
\end{abstract}

% REQUIRED
\begin{keywords}
  Symplectic model reduction, Hamiltonian system, Greedy basis generation, Symplectic Discrete Empirical Interpolation (SDEIM)
\end{keywords}

% REQUIRED
\begin{AMS}

\end{AMS}

@@

\appendix
\input{./chapters/appendix.tex}

\bibliographystyle{siamplain}
\bibliography{ref}
\end{document}

### 2016-11-15 greedy_hamiltonian_revised/chapters/1.introduction.tex
fist commit

Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}. These methods contain two stages: the offline and the online. In the offline stage, one explors the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation. In this stage, evaluation of the solution of the original model for multiple parameter values is required. The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.

Convectional reduced basis techniques e.g. proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space. This leads to a very expensive and often impractical offline stages when dealing with multi-dimensional paramtere domains. On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively often guided by an error estimate to certify the accuracy of the basis. This approach, accompanied with an efficient sampling procedure, trades considerable amount of computation with overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.

Beside computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, stability of the original model. In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve geometric structure. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-ordered configuration space. In this way the geometric structure of the original system is inherited by the reduced system. For Hamiltonian systems Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result the reduced system preserves the symplectic structure. Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well suitable for problems with a high-dimensional parameter domain.

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems, as an iterative routine that successively enriches the reduced basis at each iteration. On one hand this technique results in a reduced Hamiltonian system that mimics symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time integration. On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches. Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm. For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.

>>>>

Reduced basis methods have emerged as a powerful approach for the reduction of the intrinsic complexity of such models \cite{Ito:1998up,Ito:1998ch,Ito:2001ev,Peterson:1989ki}. These methods contain two stages: {\edit the offline stage and the online stage}. In the offline stage, one explors the parameter space to construct a low-dimensional basis that accurately represents the parametrized solution to the partial differential equation. In this stage, evaluation of the solution of the original model for multiple parameter values is required. The online stage comprises a Galerkin projection onto the span of the reduced basis, which allows exploration of the parameter space at significantly reduced complexity \cite{Antoulas:2005:ALD:1088857,Anonymous:2016wl}.

Convectional reduced basis techniques e.g. proper orthogonal decomposition (POD) \cite{Kunisch:2002er,Atwell:2001ja,Ravindran:2002hn}, require the exploration of the entire parameter space. This leads to a very expensive and often impractical offline {\edit stage} when dealing with multi-dimensional paramtere domains. On the other hand, sampling techniques, usually of a greedy nature, search through the parameter space selectively often guided by an error estimate to certify the accuracy of the basis. This approach, accompanied with an efficient sampling procedure, trades considerable amount of computation with overall accuracy of the reduced-basis \cite{Cuong:2005gd,Rozza:2005ie,Anonymous:2016wl}.

{\edit Besides} computational complexity, another aspect of reduced order modeling is the preservation of structures and, in particular, {\edit the} stability of the original model. In general, reduced order models do not guarantee that such properties are preserved \cite{Anonymous:pMn0O0Q4}.

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve geometric structure. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-ordered configuration space. In this way the geometric structure of the original system is inherited by the reduced system. {\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al. \cite{Chaturantabut:2016he}, Polyuga et al. \cite{Polyuga:2010gj} and the references therein. These works construct a reduced port-Hamiltonian system using Krylov or POD methods that carry the passivity and stability of the original system.} For Hamiltonian systems Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result the reduced system preserves the symplectic structure. Although these methods preserve geometric structure, they use a POD-like approach for constructing the reduced basis, and are not well {\edit suited} for problems with a high-dimensional parameter domain.

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems, as an iterative routine that successively enriches the reduced basis at each iteration. On one hand this technique results in a reduced Hamiltonian system that mimics symplectic properties of the original system and preserves the Hamiltonian structure and stability over the course of time integration. On the other hand, since time integration of the original system is only required once per iteration, the proposed method saves substantial computational cost during the offline stage when compared to alternative POD-like approaches. {\edit It is well known that structured matrices, e.g. symplectic matrices, are not generally well-conditioned \cite{Karow:2006cf}. The greedy update of the symplectic basis presented here, yields a orthosymplectic basis, and therefore a norm bounded basis.} Moreover, we demonstrate that assumptions, natural for the set of all solutions of the original Hamiltonian system under variation of parameters, lead to exponentially fast convergence of the greedy algorithm. For nonlinear Hamiltonian systems, we show how the basis can be combined with the discrete empirical interpolation method (DEIM) {\edit \cite{Chaturantabut:2010cz,Barrault:2004kz}} to allow a fast evaluation of nonlinear terms while maintaining the symplectic structure.

### 2016-11-15 greedy_hamiltonian_revised/chapters/2.mor.tex
fist commit

We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of parameter $\omega$
\begin{equation} \label{eq:MoOr:2}
	\mathcal M = \{ \mathbf x(\omega) | \omega \in \Gamma \} \subset \mathbb R^n.
\end{equation}
Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ with arbitrary accuracy. By abuse of notation we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.

@@

\subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}). If we Suppose that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$. The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be minimum. The POD basis of size $k$ is thus the solution of the optimization problem
\begin{equation} \label{eq:MoOr:6}
\begin{aligned}
& \underset{V\in \mathbb R^{n\times k}}{\text{minimize}}
& & \| S - VV^TS\|_F \\
& \text{subject to}
& & V^TV = I_k
\end{aligned}
\end{equation}
Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$. According to Schmidt-Mirsky-Eckard-Young theorem \cite{Antoulas:2005:ALD:1088857}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
\begin{equation} \label{eq:MoOr:7}
	V = \sigma_1 u_1 v^T_1 + \cdots + \sigma_k u_k v^T_k.
\end{equation}
Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively.

\subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
Let us now discuss the inefficiency of evaluating nonlinearities when using projection based reduced model. Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function. Now suppose a $k$-dimensional reduced basis $V$ is provided. The reduced system takes the form
\begin{equation} \label{eq:MoOr:8}
	\frac{d}{dt} \mathbf y = \underbrace{(WV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(WV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
\end{equation}
Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system. However, evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system. Suppose that evaluation of $\mathbf g$ with $n$ components has complexity $\alpha(n)$, for some function $\alpha$. Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the evaluation of the nonlinear function. This means that estimation of the nonlinear terms can be as expensive as solving the original system.

To overcome this bottleneck we take an approach similar to what we did above. Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
\begin{equation} \label{eq:MoOr:10}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega).
\end{equation}
Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the corresponding vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
\begin{equation} \label{eq:MoOr:11}
	P = [e_{p_1},\dots,e_{p_m}],
\end{equation}
where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$. Hence, multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$. If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
\begin{equation} \label{eq:MoOr:12}
	P^T \mathbf g = (P^TU)\mathbf c.
\end{equation}
Finally the approximation of $\mathbf g$ is determined by
\begin{equation} \label{eq:MoOr:13}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega) = U (P^TU)^{-1} P^T \mathbf g(t,\mathbf x,\omega),
\end{equation}
which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM on the reduced system (\ref{eq:MoOr:5}) yields
\begin{equation} \label{eq:MoOr:14}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(t,V\mathbf y , \omega).
\end{equation}
Note that the matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and since $\mathbf g$ is evaluated only at $m$ of its components, the evaluation of the nonlinear term in (\ref{eq:MoOr:14}) does not depend on $n$.

To obtain the projection basis $U$, the POD can be applied to the ensemble of the discrete evaluation of $\mathbf g$. There is no additional cost for computing the nonlinear snapshots since they are generated when computing the trajectory snapshot matrix $S$. The interpolating indices $p_1,\dots,p_m$ can be constructed as follows. Given the projection basis $U = \{u_1,\dots,u_m\}$, the first interpolation index $p_1$ is chosen according to the component of $u_1$ with the largest magnitude. The rest of the interpolation indices, $p_2,\dots,p_m$ correspond to the component of the largest magnitude of the residual vector $\mathbf r = u_l - U \mathbf c$. It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined. We summarize the algorithm of generating interpolating indices in Algorithm \ref{alg:MoOr:1}.

@@

If we use an implicit time integration scheme, a similar inefficiency occurs for the computation of Newton iteration. In addition to the nonlinear term, the Jacobian of $\mathbf g(\mathbf x)$ with respect to the reduced state vector must be computed at each iteration. The Jacobian takes the form
\begin{equation} \label{eq:MoOr:9}
	\mathbf J_{\mathbf y}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf x}(\mathbf g) V.
\end{equation}
The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system. Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent. Very often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
\begin{equation} \label{eq:MoOr:15}
	\mathbf g(\mathbf x) =
	\begin{pmatrix}
		g_1(x_1,\dots,x_n) \\
		g_2(x_1,\dots,x_n) \\
		\vdots \\
		g_n(x_1,\dots,x_n)
	\end{pmatrix}
	=
	\begin{pmatrix}
		g_1(x_1) \\
		g_2(x_2) \\
		\vdots \\
		g_n(x_n)
	\end{pmatrix}.
\end{equation}
with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
\begin{equation} \label{eq:MoOr:16}
	\tilde N(\mathbf y) \approx (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(V\mathbf y) = (WV)^{-1} U(P^TU)^{-1}\mathbf g(P^TV\mathbf y)
\end{equation}
If we now take the Jacobian of the approximate function we recover
\begin{equation}
	\mathbf J_{\mathbf y}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf x}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{k\times k}.
\end{equation}
The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only at $m\times m$ components. Hence the overall complexity of computing the Jacobian is now independent of $n$.

>>>>

We define the solution manifold as the set of all solutions to (\ref{eq:MoOr:1}) under variation of {\edit the parameters in $\Gamma$}
\begin{equation} \label{eq:MoOr:2}
	\mathcal M = \{ \mathbf x(\omega) | \omega \in \Gamma \} \subset \mathbb R^n.
\end{equation}
Note that the exact solution and solution manifold is not always available; we assume that we have a numerical integrator in hand that can approximate the solution to (\ref{eq:MoOr:1}) for any realization of $\omega$ {\edit with a given accuracy}. By abuse of notation we refer to $\mathbf x$ and $\mathcal M$ as the exact solution and the exact solution manifold, respectively, rather than the discrete solution and discrete solution manifold.

@@

\subsection{Proper Orthogonal Decomposition} \label{chap:MoOr.PrOr:1}
Let $\mathbf x (t_i,\omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$ be a finite number of samples, referred to as \emph{snapshots}, from the solution manifold (\ref{eq:MoOr:2}). If we {\edit suppose} that a reduced basis $V$ is provided, the projection operator from $\mathbb R^n$ onto the reduced space can be constructed as $VV^T$. The proper orthogonal decomposition (POD) requires the total error of projecting all the snapshots onto the reduced space to be minimum. The POD basis of size $k$ is thus the solution of the optimization problem
\begin{equation} \label{eq:MoOr:6}
\begin{aligned}
& \underset{V\in \mathbb R^{n\times k}}{\text{minimize}}
& & \| S - VV^TS\|_F \\
& \text{subject to}
& & V^TV = I_k
\end{aligned}
\end{equation}
Here $S$ is the snapshot matrix, containing snapshots $\mathbf x(t_i,\omega_j)$ in its columns, $\|\cdot \|_F$ is the Frobenius norm of a matrix and $I_k$ is the identity matrix of size $k$. According to {\edit Schmidt-Mirsky-Eckart-Young theorem \cite{Markovsky:2011:LRA:2103589}}, the solution to the optimization (\ref{eq:MoOr:6}) is equivalent to the truncated singular value decomposition (SVD) of the snapshot matrix $S$ given by
\begin{equation} \label{eq:MoOr:7}
	V = \sigma_1 u_1 v^T_1 + \cdots + \sigma_k u_k v^T_k.
\end{equation}
Here $\sigma_i, u_i$ and $v_i$ are the singular values, left singular vectors and right singular vectors of $S$, respectively {\edit \cite{Markovsky:2011:LRA:2103589} }.

\subsection{Discrete Empirical Interpolation Method (DEIM)} \label{chap:MoOr.DEIM:1} \nocite{Chaturantabut:2010cz}
In this section we {\edit discuss the efficiency} of evaluating nonlinearities when using projection based reduced model. Suppose that the right hand side function in (\ref{eq:MoOr:1}) is of the form $\mathbf f(t,\mathbf x , \omega) = L\mathbf x + \mathbf g(t,\mathbf x ,\omega)$, where $L\in \mathbb R^{n\times n}$ reflects the linear part, and $\mathbf g$ is a nonlinear function. Now suppose a $k$-dimensional reduced basis $V$ is provided. The reduced system takes the form
\begin{equation} \label{eq:MoOr:8}
	\frac{d}{dt} \mathbf y = \underbrace{(WV)^{-1} L V}_{\tilde L} \mathbf{y} + \underbrace{(WV)^{-1} \mathbf g(t,V\mathbf y,\omega)}_{\tilde N (\mathbf y)}.
\end{equation}
Here, $\tilde L$ is a $k\times k$ matrix which can be computed before time integration of the reduced system. However, evaluation of $\tilde N (\mathbf y)$ has a complexity that depends on $n$, the size of the original system. Suppose that evaluation of $\mathbf g$ with $n$ components has complexity $\alpha(n)$, for some function $\alpha$. Then the complexity of evaluating $\tilde N(\mathbf y)$ is $\mathcal{O}(\alpha(n) + 4nk)$ which consists of 2 matrix-vector operations and the evaluation of the nonlinear function. This means that estimation of the nonlinear terms can be as expensive as solving the original system.

{\edit To overcome this bottleneck we take an approach similar to section \ref{chap:MoOr.PrOr:1} \cite{Chaturantabut:2010cz,Barrault:2004kz}}. Assume that the manifold $\mathcal M_{\mathbf g} = \{ \mathbf g(t,\mathbf x , \omega)| t\in \mathbb R, \mathbf x \in \mathbb R , \omega \in \Gamma\}$ is of a low dimension and that $\mathbf g$ can be approximated by a linear subspace of dimension $m\ll n$, spanned by the basis $\{ u_1 , \dots , u_m \}$, i.e.
\begin{equation} \label{eq:MoOr:10}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega).
\end{equation}
Here $U$ contains basis vectors $u_i$ and $\mathbf c$ is the corresponding vector of coefficients. Now suppose $p_1,\dots,p_m$ are $m$ indices from $\{1,\dots,n\}$ and define an $n\times m$ matrix
\begin{equation} \label{eq:MoOr:11}
	P = [e_{p_1},\dots,e_{p_m}],
\end{equation}
where $e_{p_i}$ is the $p_i$-th column of the identity matrix $I_n$. Hence, multiplying $P$ with $\mathbf g$ selects components $p_1,\dots,p_m$ of $\mathbf g$. If we assume that $P^TU$ is non-singular, the coefficient vector $\mathbf c$ can be uniquely determined from
\begin{equation} \label{eq:MoOr:12}
	P^T \mathbf g = (P^TU)\mathbf c.
\end{equation}
Finally the approximation of $\mathbf g$ is determined by
\begin{equation} \label{eq:MoOr:13}
	\mathbf g(t,\mathbf x,\omega) \approx U \mathbf c(t,\mathbf x,\omega) = U (P^TU)^{-1} P^T \mathbf g(t,\mathbf x,\omega),
\end{equation}
which is referred to as the \emph{Discrete Empirical Interpolation} (DEIM) approximation \cite{Chaturantabut:2010cz}. Applying DEIM on the reduced system (\ref{eq:MoOr:5}) yields
\begin{equation} \label{eq:MoOr:14}
	\frac{d}{dt} \mathbf y = \tilde L \mathbf y + (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(t,V\mathbf y , \omega).
\end{equation}
Note that the matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and since $\mathbf g$ is evaluated only at $m$ of its components, the evaluation of the nonlinear term in (\ref{eq:MoOr:14}) does not depend on $n$.

{\edit To obtain the projection basis $U$, the POD can be applied to the \emph{nonlinear snapshots}. The nonlinear snapshots are the ensemble of samples of the nonlinear term $\mathbf g(t_i,\mathbf x, \omega_j)$ with $i=1,\dots,m$ and $j=1,\dots,n$}. There is no additional cost for computing the nonlinear snapshots, since they are generated when computing the trajectory snapshot matrix $S$. The interpolating indices $p_1,\dots,p_m$ can be constructed as follows. Given the projection basis $U = \{u_1,\dots,u_m\}$, the first interpolation index $p_1$ is chosen according to the component of $u_1$ with the largest magnitude. The rest of the interpolation indices, $p_2,\dots,p_m$ correspond to the component of the largest magnitude of the residual vector $\mathbf r = u_l - U \mathbf c$. It is shown in \cite{Chaturantabut:2010cz} that if the residual vector is a nonzero vector in each iteration then $P^TU$ is non-singular and (\ref{eq:MoOr:13}) is well defined. We summarize the algorithm of generating interpolating indices in Algorithm \ref{alg:MoOr:1}.

@@

{\edit Numerical solution of (\ref{eq:MoOr:8}) sometimes involve computation of the Jacobian of the nonlinear function $\mathbf g(t,\mathbf x, \omega)$ with respect to the reduced state variable $\mathbf y$}
\begin{equation} \label{eq:MoOr:9}
	\edit \mathbf J_{\mathbf x}(\mathbf g) = (WV)^{-1} \mathbf J_{\mathbf y}(\mathbf g) V,
\end{equation}
{\edit where $\mathbf J_\alpha(\mathbf g)$ is the Jacobian matrix of $\mathbf g$ with respect to the variable $\alpha$.} The complexity of computing (\ref{eq:MoOr:9}) is $\mathcal{O}(\alpha(n) +2n^2k+2nk^2+2nk)$ comprising several matrix-vector multiplications and an evaluation of the Jacobian which depends on the size of the original system. Approximating the Jacobian in (\ref{eq:MoOr:9}) is usually both problem and discretization dependent. Very often the nonlinear function $\mathbf g$ is evaluated component-wise i.e.
\begin{equation} \label{eq:MoOr:15}
	\mathbf g(\mathbf x) =
	\begin{pmatrix}
		g_1(x_1,\dots,x_n) \\
		g_2(x_1,\dots,x_n) \\
		\vdots \\
		g_n(x_1,\dots,x_n)
	\end{pmatrix}
	=
	\begin{pmatrix}
		g_1(x_1) \\
		g_2(x_2) \\
		\vdots \\
		g_n(x_n)
	\end{pmatrix}.
\end{equation}
with $\mathbf x$ the configuration vector. In such cases the interpolating index matrix $P$ and the nonlinear function $\mathbf g$ commute, i.e.,
\begin{equation} \label{eq:MoOr:16}
	\tilde N(\mathbf y) \approx (WV)^{-1} U(P^TU)^{-1}P^T \mathbf g(V\mathbf y) = (WV)^{-1} U(P^TU)^{-1}\mathbf g(P^TV\mathbf y)
\end{equation}
If we now take the Jacobian of the approximate function we recover
\begin{equation}
	\edit \mathbf J_{\mathbf x}(\mathbf g) = \underbrace{ (WV)^{-1} U(P^TU)^{-1} }_{k\times m} \underbrace{ \mathbf J_{\mathbf y}(\mathbf g(P^T V \mathbf y) ) }_{m\times m} \underbrace{P^T V}_{m\times k}.
\end{equation}
The matrix $(WV)^{-1} U(P^TU)^{-1}$ can be computed offline and the Jacobian is evaluated only at $m\times m$ components. Hence the overall complexity of computing the Jacobian is now independent of $n$. {\edit We refer the reader to \cite{Barrault:2004kz,Chaturantabut:2010cz} for more detail.}
